\chapter*{Publications}

\section*{International journals}

\begin{enumerate}[label={[J\arabic*]}]
    \item \myself, T. Zugno, M. Polese and M. Zorzi, ``Resource Management for 5G NR Integrated Access and Backhaul: A Semi-Centralized Approach", IEEE Transactions on Wireless Communications, vol. 21, no. 2, pp. 753-767, Feb. 2022.
    \item \textit{Early access, to appear in IEEE/ACM Transactions on Networking.} A. A. Gargari, A. Ortiz, \myself, W. de Sombre, M. Zorzi and A. Asadi, ``Risk-Averse Learning for Reliable mmWave Self-Backhauling", 2024.
    \item \textit{To appear in IEEE Transactions on Communications.} A. Rech, L. Badia, S. Tomasin, \myself, M. Giordani, J. Gambini and M. Zorzi, ``Downlink Clustering-Based Scheduling of IRS-Assisted Communications With Reconfiguration Constraints", 2024.
    \item \textit{Under review in IEEE Transactions on Wireless Communications.} M. Rawat, \myself,
    M. Giordani, L.-A. Dufrene, Q. Lampin and M. Zorzi, ``Optimizing Energy Efficiency of 5G RedCap Beam Management for Smart Agriculture Applications", 2024. 
    \item \textit{To be submitted to IEEE Transactions on Communications.} \myself, L. Badia and M. Zorzi,
    ``Interaction of Sources With Full Online Information as a Markov Game of Age of Information".
    \item \textit{To be submitted to IEEE/ACM Transactions on Networking.} \myself, A. Traspadini, M.
    Giordani and M. Zorzi, ``End-to-end Simulation of 5G NR Integrated Access and Backhaul Networks for Remote Connectivity".
\end{enumerate}

\section*{Conference proceedings}

\begin{enumerate}[label={[C\arabic*]}]
    \item \myself, S, Lagén, B. Bojovic, M. Polese and M. Zorzi, ``Improving the Efficiency of MIMO Simulations in ns-3", Proceedings of the 2023 Workshop on ns-3 (WNS3 23), Association for Computing Machinery, New York, NY, USA, 2023.
    \item \myself{} et al. ``End-to-End Simulation of 5G Networks Assisted by IRS and AF Relays", 20th Mediteranean Communication and Computer Networking Conference (MedComNet), 2022.
    \item A. A. Gargari, A. Ortiz, \myself, A. Klein, M. Hollick, M.Zorzi and A. Asadi, ``Safehaul: Risk-Averse Learning for Reliable mmWave Self-Backhauling in 6G Networks", IEEE INFOCOM, 2023.
    \item H. Poddar, T. Yoshimura, \myself, T. Rappaport, A. Ishi and M. Zorzi, ``ns-3 Implementation of Sub-Terahertz and Millimeter Wave Drop-based NYU Channel Model (NYUSIM)", Proceedings of the 2023 Workshop on ns-3 (WNS3 23), Association for Computing Machinery, New York, NY, USA, 2023.
    \item A. A. Gargari, \myself, A. Ortiz, N. M. Gholian, M. Polese and M. Zorzi, ``Demo:SeBaSi system-level Integrated Access and Backhaul simulator for self-backhauling", 2023 IEEE 24th International Symposium on a World of Wireless, Mobile and Multimedia Networks (WoWMoM), Boston, MA, USA, 2023.
    \item M. Sandri, \myself, M. Giordani and M. Zorzi, ``Implementation of a Channel Model for Non-Terrestrial Networks in ns-3", Proceedings of the 2023 Workshop on ns-3, 2023.
    \item \myself, T. Zugno, M. Giordani, L.-A. Dufrene, Q. Lampin and M. Zorzi, ``5G NR-Light
    at Millimeter Waves: Design Guidelines for Mid-Market IoT Use Cases," 2023 International
    Conference on Computing, Networking and Communications (ICNC), Honolulu, HI, USA, 2023.
    \item A. Gargari, \myself, M. Polese and M. Zorzi, ``6G Integrated Access and Backhaul
    Networks with Sub-Terahertz Links," 2023 18th Wireless On-Demand Network Systems and
    Services Conference (WONS), Madonna di Campiglio, Italy, 2023.
    \item A. Rech, \myself, S. Tomasin, F. Moretto, L. Badia, M. Giordani, J. Gambini and M. Zorzi, ``Downlink TDMA Scheduling for IRS-aided Communications with Block-Static Constraints", 2023 IEEE Wireless Communications and Networking Conference (WCNC), 2023.
    \item \myself, L. Badia and M. Zorzi, M, ``A Markov Game of Age of Information From Strategic Sources With Full Online Information", 2023 IEEE International Conference on Communications (ICC), 2023.
    \item H. Poddar, T. Yoshimura, \myself, T.T. Rappaport, A. Ishii and M. Zorzi, ``Full Stack End-To-End mmWave Simulations Using 3GPP and NYUSIM Channel Model in ns-3", IEEE International Conference on Communications (1CC), 2023.
\end{enumerate}