\clearpairofpagestyles
\rohead*{}
\rehead*{}
\lohead*{}
\lehead*{}
\rofoot{}
\lefoot{}

\small

First, I wish to thank my supervisor, Michele Zorzi, who gave me the chance of experiencing what it means to ``be a researcher", attend international conferences and collaborate with bright colleagues from diverse backgrounds.
Thanks to him, and professors Tommaso Melodia and Michele Polese, I also had the opportunity to spend seven
months as a Visiting Researcher at Northeastern University, Boston. This has probably been the highlight of my Ph.D., and I am deeply grateful for having been granted the opportunity to visit such a prestigious University.

Special thanks are also dedicated to all my co-authors, from all over the world:
Tommaso Zugno, Michele Polese, Marco Giordani, Michele Zorzi,
Alberto Rech, Stefano Tomasin, Leonardo Badia, Federico Agostini, Louis-Adriene Dufrene, Quentin Lampin, Manishika Rawat,
Amir Ashtari Gargari, Andrea Ortiz, Arash Asadi, Anja Klein, Matthias Hollick,
Mattia Sandri,
Jonathan Gambini, Federico Moretto,
Hitesh Poddar, Theodore S. Rappaport, Tomoki Yoshimura, Art Ishii,
Sandra Lagén, Biljana Bohovic.

Thanks to the ``friends of old": Furia, Federica, Sara, Mama, Anna, Matty, Dalla, Pippo, Cami, Crive, Fecchio, Olga, Cate, Giorgia. Thanks for sticking with me.

Thanks to V.B. 12 and its numerous inhabitants. A special thanks to Mama, Sara and Mike.

I also wish to express my gratitude to the wonderful people that I met during this journey.
You are arguably the greatest gift that the Ph.D. provided me. 
Thanks, Alberto (Jesus) Z., Alberto R., Alessia P., Donald S., Giulia R., Laura C., Daniele M., Matteo D., Paolo T., Mattia L., Ingrid H., Gianmaria V., Eleonora C., Alessandro T., Pietro T., Aurora P., Sara L., Federica B, Gabriele R.
Thanks to all people of SIGNET, 219, LTTM, Mian and Sgabuzzino.
Thanks to Civico 4 for fueling our dreams.
I hope I did not forget anyone. In case I did, sorry, but it's late night.

Thanks to whoever made my time in the U.S. so special. Thanks to Michele P., Salvatore D'O., Leonardo B. and Stefano M. for welcoming me.
Thanks to all the people I shared wonderful memories with: Tommaso R., Raoul R., Marco C., Ivan P., Antonio B., Nicola Z., Marco G., Grimo, Pietro S., Andrea L., Samuel Q., Kristen L..
Thanks to the ``The Bon'' and the ``Lower Level'' for hosting the best U.S. memories.

Thanks to the people I met during the Lipari Summer School and the ARQUS Training Week in Wroclaw.

Grazie alla mia famiglia, senza la quale nulla di tutto questo sarebbe stato possibile.

Infine, un grazie speciale a Chiara. Sei, e sei stata (con qualche cambio di ruolo) un dono meraviglioso che questo percorso mi ha regalato.

\normalsize
Thank you all for having shared this journey with me.